The Precambrian paleogeography of Laurentia

### 2021-07-23 Manuscript/Laurentia_Paleogeo_Manuscript.tex
edits associated with corrections of chapter proofs

\begin{abstract}
Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its amalgamation 1.8 billion years ago. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic to Mesoproterozoic supercontinent Nuna and latest Mesoproterozoic to Neoproterozoic supercontinent Rodinia. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years. 
\end{abstract}
%\keywords{First keyword \and Second keyword \and More} % (optional)
\textit{This manuscript is a preprint of the chapter: \vspace{0.1 cm} \\
Swanson-Hysell, N. L. (2021) The Precambrian paleogeography of Laurentia. In: Pesonen, L.J., Salminen, J., Evans, D.A.D., Elming, S.-Å., Veikkolainen, T. (eds.) Ancient Supercontinents and the Paleogeography of the Earth.}
\vspace{0.3 cm}
  \end{@twocolumnfalse} % Method A for two-column formatting
] % Method A for two-column formatting

@@

As for the suturing of the Wyoming province to Laurentia (Fig. \ref{fig:Laurentia_map}), many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 Ga tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, evidence for earlier orogenesis ca. 1.78 to 1.75 Ga in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between the Wyoming and Superior provinces. Regardless, in both of these interpretations Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast.

These collisional orogenies, particularly the Trans-Hudson orogeny, are interpreted to be associated with assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}. The Trans-Hudson orogeny is taken to be the terminal collision associated with closure of the Manikewan Ocean, a large oceanic tract that had previously separated the Superior and the North Atlantic provinces from the composite Slave+Rae+Hearne provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}; Fig. \ref{fig:Superior_Slave_recons}). The paleogeographic model of \cite{Pehrsson2015a} posits the closure of the Manikewan Ocean not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia. In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography. In their model,  Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}). This core then subsequently grows to be a semi-supercontinent with India and Australia; however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.

\begin{figure*}
\centering
\includegraphics[width=5.25 in]{../Figures/Fig2_Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model follows \cite{Swanson-Hysell2021a} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 20 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis. In the northwest, this accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margin of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the southern margin of the North Atlantic province (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}). The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a, Swanson-Hysell2021a}.

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwestern Laurentia to the Ogilvie Mountains of northwestern Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g. \citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development. Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in southwestern Laurentia \citep{Yonkee2014a, Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting off the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia,  it suggests that significant lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up. One explanation is that Tonian extensional basin development was associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence, there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

@@

There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenville orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands for which the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained are included (Table 2). Additional not-rated poles, published after the ratings were conferred, that are included in the present compilation are: 1) a pole from the ca. 1780 Ma East Central Minnesota Batholith that confirms that the Slave, Hearne, Rae, and Superior provinces were part of coherent amalgamated Laurentia following the Trans-Hudson orogeny \citep{Swanson-Hysell2021a}; 2) a pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at that time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. These poles will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2020a}, incorporating data from \cite{Weil2004a}, are also included.

@@

There are poles in the compilation for the Slave, Wyoming, Rae, Superior and North Atlantic provinces prior to Laurentia amalgamation (Fig. \ref{fig:Laurentia_poles} and Table 2). Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. A lingering question raised in \citet{Hoffman1988a} is to what extent the Archean provinces each had independent drift histories with significant separation or shared histories before experiencing fragmentation and reamalgamation. The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}. High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their present-day relative orientation to one another and having distinct pole paths as constrained by five time slices of nearly coeval poles between 2.23 and 1.89 Ga (Fig. \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}). These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion. The data also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of the Manikewan Ocean between the Superior province and the Hearne+Rae+Slave provinces. Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.

@@

\begin{table}[hbt]
\caption{Rotations of separated terranes}
{\scriptsize
\begin{tabular}{|l|l|l|l|p{1.1 in}|}
  \hline
& Euler pole & Euler pole & rotation & note and \\
Terrane & longitude & latitude & angle & citation \\
\hline
Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
\hline
Scotland & 161.9 & 78.6 & -31.0 & Reconstruction of Atlantic Ocean opening following \cite{Torsvik2017a} \\
\hline
Svalbard & 305.0 & 81.0 & -68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
\hline
\end{tabular}
}
\label{tab:terrane_rotations}
\end{table}

Through the Proterozoic, there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces (LIPs) can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt-Purcell Supergroup and the ca. 1.1 Ga Midcontinent Rift (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2020a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths (Fig. \ref{fig:Laurentia_poles}) such as ascending to the ca. 1140 to 1108 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca. 1108 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca. 980 Ma Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2020a}.

Data from other terranes add resolution to the record. In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire craton. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia (\citealp{Irving1972a}; Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dikes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dike \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with older and younger constraints from mainland Laurentia while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).

@@

There are more published models that show snapshots and publish rotation parameters associated with particular time intervals, such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, without providing parameters for a continuous model. The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record. The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (although the model parameters were not published with that study they have now been made available by the authors). This paleogeographic model implements large oscillations including ones in the Ediacaran that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again. The rates of Laurentia's motion associated with these models are also summarized in Figure \ref{fig:Laurentia_paleolatitude}. Over much of the record, Laurentia's pole positions can be satisfied through motion of the continent at rates of $<$10 cm/yr with intervals of more rapid motion such as during the Keweenawan Track and in the Paleozoic (Fig. \ref{fig:Laurentia_paleolatitude}).

>>>>

\begin{abstract}
Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its amalgamation 1.8 billion years ago. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic to Mesoproterozoic supercontinent Nuna and latest Mesoproterozoic to Neoproterozoic supercontinent Rodinia. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years. 
\end{abstract}
%\keywords{First keyword \and Second keyword \and More} % (optional)
\textit{This manuscript is a preprint of the chapter: \vspace{0.1 cm} \\
Swanson-Hysell, N. L. (2021) The Precambrian paleogeography of Laurentia. In: Pesonen, L.J., Salminen, J., Evans, D.A.D., Elming, S.-Å., Veikkolainen, T. (eds.) Ancient Supercontinents and the Paleogeography of the Earth, doi:10.1016/B978-0-12-818533-9.00009-6.}
\vspace{0.3 cm}
  \end{@twocolumnfalse} % Method A for two-column formatting
] % Method A for two-column formatting

@@

As for the suturing of the Wyoming province to Laurentia (Fig. \ref{fig:Laurentia_map}), many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 Ga tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, evidence for earlier orogenesis ca. 1.78 to 1.75 Ga in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between the Wyoming and Superior provinces. Regardless, in both of these interpretations Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast.

These collisional orogenies, particularly the Trans-Hudson orogeny, are interpreted to be associated with assembly of the supercontinent Nuna \citep{Zhang2012a,Pehrsson2015a}. The Trans-Hudson orogeny is taken to be the terminal collision associated with closure of the Manikewan Ocean, a large oceanic tract that had previously separated the Superior and the North Atlantic provinces from the composite Slave+Rae+Hearne provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}; Fig. \ref{fig:Superior_Slave_recons}). The paleogeographic model of \cite{Pehrsson2015a} posits that the closure of the Manikewan Ocean not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia. In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography. In their model,  Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}). This core then subsequently grows to be a semi-supercontinent with India and Australia; however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.

\begin{figure*}
\centering
\includegraphics[width=5.25 in]{../Figures/Fig2_Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model follows \cite{Swanson-Hysell2021b} and reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 20 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis. In the northwest, this accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was accretionary orogenesis on the southern margin of the Superior province (the Penokean orogeny; \citealp{Schulz2007a}) and the southern margin of the North Atlantic province (the Makkovik-Ketilidian orogeny; \citealp{Kerr1996a}). The Penokean orogeny involved the accretion of a microcontinent block (the Marshfield terrane) and arc terranes to the southern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a, Swanson-Hysell2021b}.

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwestern Laurentia to the Ogilvie Mountains of northwestern Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g. \citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development. Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in southwestern Laurentia \citep{Yonkee2014a, Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting off the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia,  it suggests that significant lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up. One explanation is that Tonian extensional basin development was associated with transtension in a dominantly strike-slip tectonic regime \citep{Smith2015b, Strauss2015a}. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence, there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of the tectonic scenarios proposed by \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that dominantly occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence (Fig. \ref{fig:.

@@

There are similar issues with the rich data sets associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. The ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenville orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citealp{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands for which the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained are included (Table 2). Additional not-rated poles, published after the ratings were conferred, that are included in the present compilation are: 1) a pole from the ca. 1780 Ma East Central Minnesota Batholith that confirms that the Slave, Hearne, Rae, and Superior provinces were part of coherent amalgamated Laurentia following the Trans-Hudson orogeny \citep{Swanson-Hysell2021b}; 2) a pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at that time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. These poles will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwestern Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2020a}, incorporating data from \cite{Weil2004a}, are also included.

@@

There are poles in the compilation for the Slave, Wyoming, Rae, Superior and North Atlantic provinces prior to Laurentia amalgamation (Fig. \ref{fig:Laurentia_poles} and Table 2). Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. A lingering question raised in \citet{Hoffman1988a} is to what extent the Archean provinces each had independent drift histories with significant separation or shared histories before experiencing fragmentation and reamalgamation. The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}. High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their present-day relative orientation to one another and having distinct pole paths as constrained by five time slices of nearly coeval poles between 2.23 and 1.89 Ga (Fig. \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}). These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion. The data also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of the Manikewan Ocean between the Superior province and the Hearne+Rae+Slave provinces. Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the differences in implied orientation and paleolatitude that result from these well-constrained poles.

@@

\begin{table}[hbt]
\caption{Rotations of separated terranes}
{\scriptsize
\begin{tabular}{|l|l|l|l|p{1.1 in}|}
  \hline
& Euler pole & Euler pole & rotation & note and \\
Terrane & longitude & latitude & angle & citation \\
\hline
Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
\hline
Scotland & 161.9 & 78.6 & -31.0 & Reconstruction of Atlantic Ocean opening following \cite{Torsvik2017a} \\
\hline
Svalbard & 305.0 & 81.0 & -68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin correlation following to \cite{Maloof2006a}\\
\hline
\end{tabular}
}
\label{tab:terrane_rotations}
\end{table}

Through the Proterozoic there are intervals where there are abundant paleomagnetic poles that constrain Laurentia's position and intervals when the record is sparse (shown colored by age in Fig. \ref{fig:Laurentia_poles}). To further visualize the temporal coverage of the poles and to summarize the motion, implied paleolatitudes for an interior point on Laurentia are shown in Figure \ref{fig:Laurentia_paleolatitude}. The ages of the utilized paleomagnetic poles are also shown in comparison to the simplified summary of tectonic events in Figure \ref{fig:tectonic_history}. Both collisional and extensional tectonism can result in the formation of lithologies that can be used to develop paleomagnetic poles either as a result of basin formation, magmatism or both. In addition, intraplate magmatism resulting from plume-related large-igneous provinces (LIPs) can lead to paleomagnetic poles in periods that are otherwise characterized by tectonic quiescence (e.g. the ca. 1267 Ma Mackenzie LIP; Fig. \ref{fig:tectonic_history}). Intracontinental rifts have led to the highest density of poles both in the case of the ca. 1.4 Ga Belt-Purcell Supergroup and the ca. 1.1 Ga Midcontinent Rift (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). The quality and resolution of the record from the Midcontinent Rift is aided by the voluminous magmatism that occurred in conjunction with basin formation that enables the development of a well-calibrated apparent polar wander path \citep{Swanson-Hysell2019a}. The late Tonian Period also has a number of poles including the Gunbarrel LIP (ca. 780 Ma) and Franklin LIP (ca. 720 Ma), as well as similarly-aged sedimentary rocks from western Laurentia basins \citep{Eyster2020a}. Overall, there is internal consistency among the paleomagnetic poles within intervals for which there is high-resolution coverage. These data result in progressive paths (Fig. \ref{fig:Laurentia_poles}) such as ascending to the ca. 1140 to 1108 Ma apex of the Logan Loop \citep{Robertson1971a}, down the ca. 1108 to 1080 Ma Keweenawan Track \citep{Swanson-Hysell2019a} to the ca. 980 Ma Grenville Loop \citep{McWilliams1975a} prior to a temporal gap before the late Tonian (ca. 775 to 720 Ma) path \citep{Eyster2020a}.

Data from other terranes add resolution to the record. In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once they have been rotated these poles can be used for reconstruction of the entire craton. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia (\citealp{Irving1972a}; Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dikes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dike \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with older and younger constraints from mainland Laurentia while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).

@@

There are more published models that show snapshots and publish rotation parameters associated with particular time intervals, such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, without providing parameters for a continuous model. The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record. The figure also shows the continuous implied position of Laurentia from the late Mesoproterozoic into the early Paleozoic from the model of \citet{Li2013a} (although the model parameters were not published with that study they have now been made available by the authors). This paleogeographic model implements large oscillations, including ones in the Ediacaran, that result from an interpretation that steep inclinations are the result of rapid motion of Laurentia from low to high latitudes and back again. The rates of Laurentia's motion associated with these models are also summarized in Figure \ref{fig:Laurentia_paleolatitude}. Over much of the record, Laurentia's pole positions can be satisfied through motion of the continent at rates of $<$10 cm/yr with intervals of more rapid motion such as during the Keweenawan Track and in the Paleozoic (Fig. \ref{fig:Laurentia_paleolatitude}).
